---
title: "Personality Traits and Scientific Reasoning"
author: "Moin Syed"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

```

```{r load packages, include=FALSE}

library(groundhog)
library(dplyr)
library(labelled)
library(ggplot2)
library(knitr)

```

```{r data setup, include=FALSE}
# setwd("~/traits-reasoning/Data")
dat <- read.csv("../Data/traits_reasoning_data_2025-10-24.csv")
names(dat)

```

## Abstract 

Personality traits have been shown to be related to many aspects of life. But what about scientific reasoning? We don't really know how these are related. The current study consists of an analysis of `r nrow(dat)` U.S. college students enrolled in STEM majors who completed measures of personality traits and scientific reasoning. The results indicated a lot of variability in scientific reasoning. 

## Background

Personality traits have been shown to be related to many aspects of life (Ozer & Benet-Martinez, 2006). But what about scientific reasoning? We don't really know how these are related, but it seems like finding out would be worthwhile. This is not just because we don't know--there are many questions for which we have no answers, and that is probably for a good reason. Not all questions are good or useful! Good to keep in mind. 

But here I think we are dealing with a good question. Personality traits correspond to relatively stable patterns of individual differences in thoughts, emotions, and behaviors. Variations in these individual differences have been linked to many life outcomes, including academic achievement. Scientifc reasoning is important not only for those participating in science, but also for society at large. Knowing more about how personality traits are related to scientific reasoning could help us better understand who tends to excel in this area, and thus could help align people's careers with their personalities, but also would open up new possibilities for how to tailor our approach to teaching scientific reasoning. 

### The Present Study

The purpose of the present study was to examine how personality traits are related to scientific reasoning. This was an exploratory correlational study with no *a priori* hypotheses.

## Method

The current study was **NOT** preregistered. Data and code are available at <https://github.com/syeducation/traits-reasoning>.
You can also access the data and code [by clicking on this text here](https://github.com/syeducation/traits-reasoning). 

### Participants and Procedure

```{r demographics, include=FALSE}

# create objects for mean and sd of age, to be inserted in text

age_mean <- mean(dat$age)
age_sd <- sd(dat$age)

```

The total sample in the current study consists of `r nrow(dat)` students enrolled in one of the three STEM-focused colleges at a large public university in the U.S. Midwest (*M* age = `r round(age_mean, digits = 2)`, *SD* = `r round(age_sd, digits = 2)`). Participants were recruited from a list of all first-year students in the three colleges who identified as racial/ethnic minorities. Eligible students were sent a survey link via email and compensated $25 for their participation. 

### Measures

#### Personality Traits.

Participants completed the 100-item Big Five Aspect Scale (DeYoung et al., 2007), which assesses the big five traits as well as ten aspects. We collected these, but we aren't using them in the current study (despite the title of the project). 

#### Scientific Reasoning.

```{r scientific reasoning, include=FALSE}

# first need to recode the scale ao that 1 always indicates a correct answer

dat <- dat %>% mutate(srs_1_scored = recode(srs_1, `1` = 1, `2` = 0),
                      srs_2_scored = recode(srs_2, `1` = 0, `2` = 1),
                      srs_3_scored = recode(srs_3, `1` = 1, `2` = 0),
                      srs_4_scored = recode(srs_4, `1` = 0, `2` = 1),
                      srs_5_scored = recode(srs_5, `1` = 0, `2` = 1),
                      srs_6_scored = recode(srs_6, `1` = 0, `2` = 1),
                      srs_7_scored = recode(srs_7, `1` = 1, `2` = 0),
                      srs_8_scored = recode(srs_8, `1` = 0, `2` = 1),
                      srs_9_scored = recode(srs_9, `1` = 0, `2` = 1),
                      srs_10_scored = recode(srs_10, `1` = 0, `2` = 1),
                      srs_11_scored = recode(srs_11, `1` = 0, `2` = 1))

# pull these together into an object for further use

srs_items <- c("srs_1_scored", "srs_2_scored", "srs_3_scored", "srs_4_scored", "srs_5_scored", "srs_6_scored", "srs_7_scored", "srs_8_scored", "srs_9_scored", "srs_10_scored", "srs_11_scored")

# creating a new variable that is the mean score across items

dat <- dat %>% dplyr::mutate(srs = rowMeans(dat[ ,srs_items], na.rm = TRUE))

# give that new variable a label

labelled::var_label(dat$srs) <- "Scientific Reasoning Scale - Mean"

# check that all looks ok

psych::describe(dat$srs)

# get cronbach's alpha for the scale, assign to object

alpha_srs <-  psych::alpha(dplyr::select(dat, "srs_1_scored", "srs_2_scored", "srs_3_scored", "srs_4_scored", "srs_5_scored", "srs_6_scored", "srs_7_scored", "srs_8_scored", "srs_9_scored", "srs_10_scored", "srs_11_scored"))

# pull out just the alpha and save as object to insert into text

srs_alpha <- alpha_srs$total$std.alpha

# create object for mean and sd for full scale to insert into text

srs_mean <- mean(dat$srs)
srs_sd <- sd(dat$srs)

# create object for mean and sd for each item to insert into table

srs_1_mean <- mean(dat$srs_1_scored)
srs_2_mean <- mean(dat$srs_2_scored)
srs_3_mean <- mean(dat$srs_3_scored)
srs_4_mean <- mean(dat$srs_4_scored)
srs_5_mean <- mean(dat$srs_5_scored)
srs_6_mean <- mean(dat$srs_6_scored)
srs_7_mean <- mean(dat$srs_7_scored)
srs_8_mean <- mean(dat$srs_8_scored)
srs_9_mean <- mean(dat$srs_9_scored)
srs_10_mean <- mean(dat$srs_10_scored)
srs_11_mean <- mean(dat$srs_11_scored)
srs_1_sd <- sd(dat$srs_1_scored)
srs_2_sd <- sd(dat$srs_2_scored)
srs_3_sd <- sd(dat$srs_3_scored)
srs_4_sd <- sd(dat$srs_4_scored)
srs_5_sd <- sd(dat$srs_5_scored)
srs_6_sd <- sd(dat$srs_6_scored)
srs_7_sd <- sd(dat$srs_7_scored)
srs_8_sd <- sd(dat$srs_8_scored)
srs_9_sd <- sd(dat$srs_9_scored)
srs_10_sd <- sd(dat$srs_10_scored)
srs_11_sd <- sd(dat$srs_11_scored)

# label each item in scale....may or may not use, but good to have

labelled::var_label(dat$srs_1_scored) <- "In a taste test, a researcher puts Brand A coffee in a cup with white tape on it and Brand B coffee in an identical cup with black tape on it. A lab assistant gives tasters one of the cups, while the researcher watches their facial expressions.
True or False? The lab assistant should not watch the cups being filled."
labelled::var_label(dat$srs_2_scored) <- "A researcher finds that American states with larger parks have fewer endangered species.
True or False? These data show that increasing the size of American state parks will reduce the number of endangered species."
labelled::var_label(dat$srs_3_scored) <- "A researcher has subjects put together a jigsaw puzzle either in a cold room with a loud radio or in a warm room with no radio. Subjects solve the puzzle more quickly in the warm room with no radio. 
True or False? The scientist cannot tell if the radio caused subjects to solve the puzzle more slowly."
labelled::var_label(dat$srs_4_scored) <- "An education researcher wants to measure the general math ability of a sample of high-performing math students. All the students have taken classes in geometry and pre-calculus.
True or False? The education researcher can measure general math ability by giving the students a geometry test."
labelled::var_label(dat$srs_5_scored) <- "Two scientists test an anti-acne cream on teenagers with acne. Scientist A wants to give the cream to all the teenagers in the study. Scientist B wants to give the cream to half the teenagers and give a cream without anti-acne ingredients to the other half.
True or False? Both ways of testing the cream are equally good."
labelled::var_label(dat$srs_6_scored) <- "A researcher has a group of subjects play a competitive game. Each subject’s goal is to make money by buying and selling tokens. Subjects are paid a flat fee for participating in the experiment.
True or False? The researcher can confidently state that the behavior in the experiment reflects real-life buying and selling behavior."
labelled::var_label(dat$srs_7_scored) <- "A randomly selected sample of Americans is surveyed about disease A before and after a 6-month media campaign about the disease. Mid-way through the media campaign, a famous celebrity dies of Disease A. The survey data indicate that knowledge of Disease A is higher after the campaign.
True or False? The media campaign may not have increased knowledge of Disease A"
labelled::var_label(dat$srs_8_scored) <- "Subjects in an experiment must press a button whenever a blue dot flashes on their computer screen. At first, the task is easy for subjects. But as they continue to perform the task, they make more and more errors.
True or False? The blue dot must flash more quickly as the task progresses."
labelled::var_label(dat$srs_9_scored) <- "Researchers want to see whether a health intervention helps school children to lose weight. School children are sorted into either an intervention or control group.
True or False? The researchers should assign the overweight children to the intervention group."
labelled::var_label(dat$srs_10_scored) <- "A researcher develops a new method for measuring the surface tension of liquids. This method is more consistent than the old method.
True or False? The new method must also be more accurate than the old method."
labelled::var_label(dat$srs_11_scored) <- "Two researchers are developing a survey to measure consumers’ feelings about customer service. Researcher A wants customers to rate their agreement with the statement “I am satisfied with customer service” on a 5-point scale, where 1 = strongly agree and 5 = strongly disagree. Researcher B wants customers to rate customer service on a 5-point scale, where 1 = not dissatisfied at all and 5 = highly dissatisfied.
True or False? These questions are equally good for measuring how consumers feel about customer service."

```

Participants completed an 11-item assessment of scientific reasoning (Drummond & Fischhoff, 2017), in which they were asked to read a description of a scientific activity and then answer True or False to a question about that activity (Cronbach's alpha = `r round(srs_alpha, digits = 2)`). 

## Results

Overall, participant did well on the scientific reasoning task, averaging more correct than incorrect answers, *M* correct = `r round(srs_mean, digits = 2)`, (*SD* = `r round(srs_sd, digits = 2)`). However, these results are best examined via tables and figures, so let's look at some.

Here is a table of each item and its rate of success:

```{r descriprtive table, echo=FALSE}

# this code creates a table, in which you almost literally "create" the table with the code
# that is, the first row (with the ~) serves as the column headers
# then the following entries indicate what should go in each cell. 

srs_table <- dplyr::tribble(
    ~Item, ~Mean, ~SD,
    "SRS Item 1", srs_1_mean,  srs_1_sd,
    "SRS Item 2", srs_2_mean,  srs_2_sd,
    "SRS Item 3", srs_3_mean,  srs_3_sd,
    "SRS Item 4", srs_4_mean,  srs_4_sd,
    "SRS Item 5", srs_5_mean,  srs_5_sd,
    "SRS Item 6", srs_6_mean,  srs_6_sd,
    "SRS Item 7", srs_7_mean,  srs_7_sd,
    "SRS Item 8", srs_8_mean,  srs_8_sd,
    "SRS Item 9", srs_9_mean,  srs_9_sd,
    "SRS Item 10", srs_10_mean,  srs_10_sd,
    "SRS Item 11", srs_11_mean,  srs_11_sd)

# now we can use the kable package to format that table

knitr::kable(srs_table, format = "simple", 
             caption = "Descriptives for SRS scale",
             align = "c", digits = 2)

```

It looks like item 11 was the most difficult: 

>`r var_label(dat$srs_11_scored)`

On average, people did pretty well, but from the standard deviation for the scale as well as the means and standard deviations of the individual items you can see there is quite a bit of variability. It is always important to plot your data, so let's take a look at the distribution!

```{r histogram, echo=FALSE}

# basic histogram code using ggplot. 
# remember for quick and dirty checks you can use the base r hist() function

ggplot(dat, aes(srs)) +
  geom_histogram(binwidth = .10, fill = "green", color = "black") +
  theme_classic() +
  labs(title = "Distribution of Scientific Reasoning Scores",
       x = "% Correct",
       y = "Frequency (Count)")
  

```

## Discussion

The purpose of the present study was to examine how personality traits are related to scientific reasoning.

Overall, it seems that people reason about science, but maybe not as much as we would have hope.

We don;t actually know how personality traits are related to scientific reasoning, because we did not assess that. That is a limitation of the study that should guide future work. 

In sum, this was a very mediocre study. We will try better in the future. 

## References

This can be done automatically, connected to Zotero, but is kind of annoying. 
