---
title: "Traits & Reasoning Project"
author: "Moin Syed"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

```

```{r load packages, include=FALSE}

library(groundhog)
groundhog.library(dplyr, "2025-04-01")
groundhog.library(psych, "2025-04-01")
groundhog.library(labelled, "2025-04-01")
groundhog.library(ggplot2, "2025-04-01")

sessionInfo()

```

```{r data setup, include=FALSE}

dat <- read.csv("Data/traits_reasoning_data_2025-10-24.csv")
names(dat)

```


## Abstract 

Blah blah blah

## Background

Personality traits are important, **so** important. 

### The Present Study

The purpose of the present study was to....

## Method

Data and Code are available at <https://github.com/syeducation/traits-reasoning>.

### Participants and Procedure

```{r demographics, include=FALSE}

age_mean <- mean(dat$age)
age_sd <- sd(dat$age)

```

The total sample in the current study consists of `r nrow(dat)` students enrolled in one of the three STEM-focused colleges at a large public university in the U.S. Midwest (*M* age = `r age_mean`, *SD* = `r age_sd`). Participants were recruited from a list of all first-year students in the three colleges who identified as racial/ethnic minorities. Eligible students were sent a survey link via email and compensated $25 for their participation. 

### Measures

#### Personality Traits.

Participants completed the 100-item Big Five Aspect Scale (BFAS), which assesses the big five traits as well as ten aspects. 

#### Scientific Reasoning.

```{r scientific reasoning, include=FALSE}

dat <- dat %>% mutate(srs_1_scored = recode(srs_1, `1` = 1, `2` = 0),
                      srs_2_scored = recode(srs_2, `1` = 0, `2` = 1),
                      srs_3_scored = recode(srs_3, `1` = 1, `2` = 0),
                      srs_4_scored = recode(srs_4, `1` = 0, `2` = 1),
                      srs_5_scored = recode(srs_5, `1` = 0, `2` = 1),
                      srs_6_scored = recode(srs_6, `1` = 0, `2` = 1),
                      srs_7_scored = recode(srs_7, `1` = 1, `2` = 0),
                      srs_8_scored = recode(srs_8, `1` = 0, `2` = 1),
                      srs_9_scored = recode(srs_9, `1` = 0, `2` = 1),
                      srs_10_scored = recode(srs_10, `1` = 0, `2` = 1),
                      srs_11_scored = recode(srs_11, `1` = 0, `2` = 1))

srs_items <- c("srs_1_scored", "srs_2_scored", "srs_3_scored", "srs_4_scored", "srs_5_scored", "srs_6_scored", "srs_7_scored", "srs_8_scored", "srs_9_scored", "srs_10_scored", "srs_11_scored")

dat <- dat %>% dplyr::mutate(srs = rowMeans(dat[ ,srs_items], na.rm = TRUE))

labelled::var_label(dat$srs) <- "Scientific Reasoning Scale - Mean"

psych::describe(dat$srs)

alpha_srs <-  psych::alpha(dplyr::select(dat, "srs_1_scored", "srs_2_scored", "srs_3_scored", "srs_4_scored", "srs_5_scored", "srs_6_scored", "srs_7_scored", "srs_8_scored", "srs_9_scored", "srs_10_scored", "srs_11_scored"))

srs_alpha <- alpha_srs$total$std.alpha

```

11-item assessment of scientific reasoning (Cronbach's alpha = `r srs_alpha`). 

## Results

```{r histogram, echo=FALSE}
hist(dat$srs)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

## Discussion

